A new tidy data structure to support exploration and modeling of temporal data

### 2019-02-03 main.Rmd
re-work section 4

The tsibble structure and operations support data pipelines for sequencing analysis. There has been a long history of pipeline discussions and implementation centering around data in various aspects. A data pipeline describes the general flow of data through an analysis, and can generally assist in conceptualizing the process as it might be applied to a variety of problems. The Extract, Transform, and Load (ETL) process from recent data warehousing literature dating back to @kimball2011data outlines the workflow to prepare data for analysis, and can be considered a data pipeline. Building a data pipeline can be technically difficult to make it sufficiently general for a variety of data, with many implementation decisions to be made about the interface, input and output objects and functionality. It is useful to articulate the data pipeline induced by new data tools.

@unix coined the term "pipelines" in software development while developing Unix at Bell Labs. In Unix-based computer operating systems, a pipeline chains together a series of operations on the basis of their standard streams, so that the output of each program becomes the input to another. This shapes the Unix toolbox philosophy: "each do one simple thing, do it well, and most importantly, work well with each other" [@unix-philosophy].

@viewing-pipeline describes a viewing pipeline for interactive statistical graphics, that takes control of the transformation from data to plot. @xgobi, @ggobi, @orca, @plumbing and @xie2014reactive implemented data pipelines for the interactive statistical software **XGobi**, **GGobi**, **Orca**, **plumbr** and **cranvas**, respectively. The pipeline is typically described with a one way flow, from data to plot. For interactive graphics, where all plots need to be updated when a user interacts with one plot, the events typically trigger the data pipeline to be run. @xie2014reactive uses a reactive programming framework, to implement the pipeline, in which user's interactions trigger a sequence of modules to update their views, that is, practically the same as running the data pipeline producing each plot.

The tidy data abstraction lays a pipeline infrastructure for data analysis modules of transformation, visualization and modeling. Each module communicates with the others, requiring tidy input, producing tidy output, chaining a series of operations together to accomplish the analytic tasks.

What is notable about an effective implementation of a data pipeline is that it coordinates a user's analysis making it cleaner to follow, and permits a wider audience to focus on the data analysis without getting lost in a jungle of computational intricacies. A fluent and fluid pipeline glues tidy data and the grammar of data manipulation together. It helps (1) break up a big problem to into manageable blocks, (2) generate human readable analysis workflow, (3) avoid introducing mistakes, at least making it possible to trace them through the pipeline.

## Time-based transformations {#sec:timepipeline}

The time-based pipeline shepherds raw temporal data through to time series analysis, and plots. It is possible and recommended to check for identical entries of key and index before analysis. Duplicates signal the data quality issue, which would likely affect succeeding analyses and hence decision making. Analysts are encouraged to gaze at data early and reason about the process of data cleaning. When the data meets the tsibble standard, it flows neatly into the analysis stage. \autoref{fig:tsibble-pipeline} illustrates the time-based pipeline.

@@

A suite of verbs are used to flatten the lumpy path from temporal data to an object that can be directly modeled in the tsibble framework, and transformed for visualization. Many time operations like lag/lead and time series models, assume an intact vector input ordered in time. To prevent inviting these errors into the analysis, it is a good practice to check and inspect any time gaps of the tsibble in the first place. Several tools are provided to understand and tackle implicit missing values in: (1) `has_gaps()` checks if there exists time gaps; (2) `scan_gaps()` reveals all implicit missing observations; (3) `count_gaps()` summarizes the time ranges that are absent from the data; (4) `fill_gaps()` turns them into explicit ones, along with imputing by values or functions.

A tsibble is an object, conceptually considered as a noun, and hence an action performed on the object can be phrased as a verb. The principle that underpins most verbs is a tsibble in and a tsibble out, thereby striving to retain a valid tsibble by automating index and key updates under the hood. Attention has been paid to error handling. If a tsibble cannot be maintained in the output of a pipeline module, for example the index is dropped by aggregation, an error informs users of the problem and suggests alternatives. This avoids surprising users and reminds them of the time context.

@@

Each verb focuses on one operation to achieve its goal. The verb should be self-explanatory to advise what it is supposed to do or fail, for example `filter()` picks observations, `select()` picks variables, and `left_join()` joins two tables. These general-purpose verbs are made available in the **tidyverse** suite. When used in the temporal context, these verbs are adapted to the time domain. A perceivable difference is summarizing variables between data frame and tsibble using `summarize()`. The former will reduce to a single summary, whereas the latter will obtain the index and their corresponding summaries. New tsibble-specific verbs (\autoref{tab:verb-table}) are provided to expand the **tidyverse** vocabulary. Users who are already familiar with **tidyverse**, will experience a gentle learning curve for mastering tsibble verbs and glide into temporal data analysis with low cognitive load.

@eopl asserted "No matter how complex and polished the individual operations are, it is often the quality of the glue that most directly determines the power of the system." Each verb works with other transformation family members in harmony. This set of verbs can result in many combinations to prepare tsibble for a broad range of visualization and modeling problems. Chaining operations is achieved with the pipe operator `%>%` introduced in the **magrittr** package [@R-magrittr], read as "then". A sequence of functions can be composed in a way that can be naturally read from left to right, which improves the readability of the code. It consequently generates a block of code without saving intermediate values. Most importantly, a new ecosystem for tidy time series analysis has been undertaken, using the tsibble framework, and is called "tidyverts", a play on tidyverse that acknowledges the time series analysis purpose.

@@

As a special case of the data frame, a tsibble pipes directly into the grammar of graphics. It is easy to create and extend specialist time-related plotting methods based on tsibble structure, such as autocorrelation plots and calendar-based graphics [@wang-calendarviz-2018].

@@

The primary force that drives the software's design choices is "data". All functions in the package **tsibble** [@R-tsibble] start with `data` or its variants as the first argument, namely "data first". This lays out a consistent interface and addresses the significance of the data throughout the software.

@@

The **purrr** package [@R-purrr] provides a good example of functional programming in R. It provides a complete and consistent set of tools to iterate each element of a vector with a function. Rolling window doesn't just iterate but rolls over a sequence of elements, with `slide()`, `tile()` and `stretch()`. `slide()` expects one input, `slide2()` two inputs, and `pslide()` multiple inputs. For type stability, the functions always return lists. Other variants including `*_lgl()`, `*_int()`, `*_dbl()`, `*_chr()` return vectors of the corresponding type, as well as `*_dfr()` and `*_dfc()` for row-binding and column-binding data frames respectively. Their multiprocessing equivalents prefixed by `future_*()` enable rolling in parallel [@R-future; @R-furrr]. This family of functions empowers users to incorporate window-related operations in their workflows.

>>>>

There has been a long history of pipeline discussions and implementation centering around data in various aspects. A data pipeline describes the general flow of data through an analysis, and can generally assist in conceptualizing the process as it might be applied to a variety of problems. @unix coined the term "pipelines" in software development while developing Unix at Bell Labs. In Unix-based computer operating systems, a pipeline chains together a series of operations on the basis of their standard streams, so that the output of each program becomes the input to another. The Extract, Transform, and Load (ETL) process from recent data warehousing literature dating back to @kimball2011data outlines the workflow to prepare data for analysis, and can also be considered a data pipeline. @viewing-pipeline describes a viewing pipeline for interactive statistical graphics, that takes control of the transformation from data to plot. @xgobi, @ggobi, @orca, @plumbing and @xie2014reactive implemented data pipelines for the interactive statistical software **XGobi**, **GGobi**, **Orca**, **plumbr** and **cranvas**, respectively. The pipeline is typically described with a one way flow, from data to plot. For interactive graphics, where all plots need to be updated when a user interacts with one plot, the events typically trigger the data pipeline to be run. @xie2014reactive uses a reactive programming framework, to implement the pipeline, in which user's interactions trigger a sequence of modules to update their views, that is, practically the same as running the data pipeline producing each plot.

Building a data pipeline can be technically difficult to make it sufficiently general for a variety of data, with many implementation decisions to be made about the interface, input and output objects and functionality. The tidy data abstraction lays a pipeline infrastructure for data analysis modules of transformation, visualization and modeling. Each module communicates with the others, requiring tidy input, producing tidy output, chaining a series of operations together to accomplish the analytic tasks.

What is notable about an effective implementation of a data pipeline is that it coordinates a user's analysis making it cleaner to follow, and permits a wider audience to focus on the data analysis without getting lost in a jungle of computational intricacies. A fluent pipeline glues tidy data and the grammar of data manipulation together. It helps (1) break up a big problem to into manageable blocks, (2) generate human readable analysis workflow, (3) avoid introducing mistakes, at least making it possible to trace them through the pipeline. New data tools developed in the R package **tsibble** [@R-tsibble] articulate the time series data pipeline, which shepherds raw temporal data through to time series analysis, and plots. More detailed explanations are given as follows.

## Time series transformation {#sec:timepipeline}

\autoref{fig:tsibble-pipeline} illustrates the distinction of a time series pipeline from a regular data pipeline. It is highly recommended to check for identical entries of key and index before constructing a tsibble. Duplicates signal the data quality issue, which would likely affect succeeding analyses and hence decision making. Analysts are encouraged to gaze at data early and reason about the process of data cleaning. When the data meets the tsibble standard, it flows neatly into the analysis stage and takes full advantage of the tsibble infrastructure.

@@

Many time operations such as lag/lead and time series models, assume an intact vector input ordered in time. Since a tsibble permits time gaps in the index, it is a good practice to check and inspect any gaps in time following the creation of a tsibble, in order to prevent inviting these avoidable errors into the analysis. The first suite of verbs (rephrasing actions performed on the object) are provided to understand and tackle implicit missing values: (1) `has_gaps()` checks if there exists time gaps; (2) `scan_gaps()` reveals all implicit missing observations; (3) `count_gaps()` summarizes the time ranges that are absent from the data; (4) `fill_gaps()` turns them into explicit ones, along with imputing by values or functions. To look into gaps over individual time periods or full-length time span, the common argument `.full` in these functions gives an option to easily switch between. The specification of `.full = TRUE` will result in fully balanced panels in other words.

@@

Besides the time gaps verbs, the **tidyverse** vocabulary is adapted and expanded to facilitate time series transformations, as listed in \autoref{tab:verb-table}. The **tidyverse** suite showcases the general-purpose verbs for effectively manipulating tabular data, for example `filter()` picks observations, `select()` picks variables, and `left_join()` joins two tables. But these verbs need handling with care when used in the time series domain. A perceivable difference is summarizing variables between data frame and tsibble using `summarize()`. The former will reduce to a single summary, whereas the latter will obtain the index and their corresponding summaries. Users who are already familiar with the **tidyverse**, will experience a gentle learning curve for mastering these verbs and glide into time series analysis with low cognitive load.

Attention has been paid to warning/error handling. The principle that underpins most verbs is a tsibble in and a tsibble out, thereby striving to maintain a valid tsibble by automatically updating index and key under the hood. If the desirable temporal ordering is changed by row-wise verbs, a warning is broadcast. If a tsibble cannot be maintained in the output of a pipeline module (likely occurring to column-wise verbs), for example the index is removed by selection, an error informs users of the problem and suggests alternatives. This avoids surprising users and reminds them of the time context.

The tsibble structure and operations support data pipelines for sequencing analysis. @eopl asserted "No matter how complex and polished the individual operations are, it is often the quality of the glue that most directly determines the power of the system." Each verb works with other transformation family members in harmony. This set of verbs can result in many combinations to prepare tsibble for a broad range of visualization and modeling problems. Chaining operations is achieved with the pipe operator `%>%` introduced in the **magrittr** package [@R-magrittr], read as "then". A sequence of functions can be composed in a way that can be naturally read from left to right, which improves the readability of the code. It consequently generates a block of code without saving intermediate values.

Most importantly, a new ecosystem for tidy time series analysis has been undertaken, using the tsibble framework, and is called "tidyverts", a play on tidyverse that acknowledges the time series analysis purpose.

@@

As a tsibble is a subclassing of data frame, it integrates well with the grammar of graphics. It is easy to create and extend specialist time series plotting methods based on the tsibble structure, for example autocorrelation plots and calendar-based graphics [@wang-calendarviz-2018].

@@

The primary force that drives the software's design choices is "data". All functions in the package **tsibble** start with `data` or its variants as the first argument, namely "data first". This lays out a consistent interface and addresses the significance of the data throughout the software.

@@

The **purrr** package [@R-purrr] provides a good example of functional programming in R. It provides a complete and consistent set of tools to iterate each element of a vector with a function. Rolling window does not just iterate but rolls over a sequence of elements, with `slide()`, `tile()` and `stretch()`. `slide()` expects one input, `slide2()` two inputs, and `pslide()` multiple inputs. For type stability, the functions always return lists. Other variants including `*_lgl()`, `*_int()`, `*_dbl()`, `*_chr()` return vectors of the corresponding type, as well as `*_dfr()` and `*_dfc()` for row-binding and column-binding data frames respectively. Their multiprocessing equivalents prefixed by `future_*()` enable rolling in parallel [@R-future; @R-furrr]. This family of functions empowers users to incorporate window-related operations in their workflows.
